\pkg{DNest4}: Diffusive Nested Sampling in \proglang{C++} and \proglang{Python}

### 2016-08-10 paper/ms.tex
Some more edits to comply with JSS

\pkg{MultiNest} \citep{feroz2009multinest} is a
\proglang{FORTRAN}
Nested Sampling implementation
that dispenses with MCMC for generating new particles from constrained
prior distributions. Instead, the new particles are generated by making
an approximation to the constrained prior using ellipsoids. This tends to work
well in low to moderate dimensional parameter spaces (up to $\sim 40$).
In higher dimensions, \pkg{DNest4} is more useful than \pkg{MultiNest}
since it is based on MCMC. We expect \pkg{MultiNest} to be more useful
than
\pkg{DNest4} only on low to moderate dimensional problems with slow
likelihood evaluations, where \pkg{DNest4}'s reliance on MCMC becomes costly.
The more recent \pkg{POLYCHORD} \citep{handley2015polychord} combines
\pkg{MultiNest}-like methods with MCMC.

@@

\subsection{Generating from the prior}
The member function used to generate straight line parameters from the
prior is:
\begin{CodeChunk}
\begin{CodeInput}
void StraightLine::from_prior(DNest4::RNG& rng)
{
   // Naive diffuse prior
   m = 1E3*rng.randn();
   b = 1E3*rng.randn();

   // Log-uniform prior
   sigma = exp(-10.0 + 20.0*rng.rand());
}
\end{CodeInput}
\end{CodeChunk}

@@

    if(which == 0)
    {
        // log_H takes care of the prior ratio
        // i.e., log_H = log(pi(theta')/pi(theta))
        log_H -= -0.5*pow(m/1E3, 2);

        // Take a step
        m += 1E3*rng.randh();

        // log_H takes care of the prior ratio
        log_H += -0.5*pow(m/1E3, 2);
    }
    else if(which == 1)
    {
        log_H -= -0.5*pow(b/1E3, 2);
        b += 1E3*rng.randh();
        log_H += -0.5*pow(b/1E3, 2);
    }
    else
    {
        // Proposal for a parameter with a log-uniform
        // prior takes the log of the parameter,
        // takes a step with respect to a uniform prior,
        // then takes the exp of the parameter
        sigma = log(sigma);
        sigma += 20.0*rng.randh();

@@

\begin{CodeChunk}
\begin{CodeInput}
m += 1E3*rng.randh();
\end{CodeInput}
\end{CodeChunk}

@@

In the example, the proposal for $\sigma$ is implemented by taking advantage
of the uniform prior for $\ln(\sigma)$. So $\sigma$ is transformed by
taking a logarithm, a proposal move is made (that satisfies detailed balance
with respect to a uniform distribution), and then $\sigma$ is exponentiated
again. The step for the uniform prior between -10 and +10
uses the following code:
\begin{CodeChunk}
\begin{CodeInput}
sigma += 20.0*rng.randh();
DNest4::wrap(sigma, -10.0, 10.0);
\end{CodeInput}
\end{CodeChunk}
The factor of 20 accounts for the prior width, and
the \code{wrap(double\&, double, double)} function uses the modulo operator
to construct periodic boundaries. For example, if the perturbation results
in a value of 10.2, which is outside the prior range, the value is modified to
$-9.8$.
The \code{wrap} function has no return value and works by modifying its
first argument, which is passed by reference.
Alternatively, the log-uniform prior, which has density proportional
to $1/\sigma$, could have been used directly by adding
$\ln\left[(1/\sigma')/(1/\sigma)\right]$ to the
return value instead of using the log/exp trick.
However, this isn't recommended for a prior distribution like this
which covers several orders of magnitude, because the appropriate scale size
for the proposal is less clear. This is likely to cause inefficient
sampling.

@@

    // Variance
    double var = sigma*sigma;

    // Conventional gaussian sampling distribution
    double log_L = 0.0;
    double mu;
    for(size_t i=0; i<y.size(); ++i)
    {
        mu = m*x[i] + b;
        log_L += -0.5*log(2*M_PI*var) - 0.5*pow(y[i] - mu, 2)/var;
    }

@@

\subsection{Parameter output}
The print and description functions are very simple:
\begin{CodeChunk}
\begin{CodeInput}
void StraightLine::print(std::ostream& out) const
{
    out<<m<<' '<<b<<' '<<sigma;
}

@@

    if(which == 0)
    {
        mu += 20.*rng.randh();
        wrap(mu, -10., 10.);
    }
    if(which == 1)
    {
        log_sigma += 20.*rng.randh();
        wrap(log_sigma, -10., 10.);
    }
    if(which == 2)
    {
        int i = rng.rand_int(n.size());
        n[i] = rng.randn();
    }

@@

    // Assemble fake dataset
    vector<double> x_fake = n;
    for(size_t i=0; i<x_fake.size(); i++)
        x_fake[i] = mu + sigma*x_fake[i];

@@

\subsection{ABC example results}
For ABC applications, there is an alternate \code{postprocess} function
called \code{postprocess_abc}.
This generates posterior samples using the $\epsilon$ value that
compresses the joint prior by 80\% as much as the final level.
\begin{CodeChunk}
\begin{CodeInput}
dnest4.postprocess_abc()
\end{CodeInput}
\end{CodeChunk}
To try different values of $\epsilon$, you can set the argument
\code{threshold\_fraction} to a value other than 0.8. For example
\begin{CodeChunk}
\begin{CodeInput}
dnest4.postprocess_abc(threshold_fraction=0.6)
\end{CodeInput}
\end{CodeChunk}

@@

    def from_prior(self):
        """
        Unlike in C++, this must *return* a numpy array of parameters.
        """
        m = 1E3*rng.randn()
        b = 1E3*rng.randn()
        sigma = np.exp(-10.0 + 20.0*rng.rand())
        return np.array([m, b, sigma])

@@

        if which == 0 or which == 1:
            logH -= -0.5*(params[which]/1E3)**2
            params[which] += 1E3*dnest4.randh()
            logH += -0.5*(params[which]/1E3)**2
        else:
            log_sigma = np.log(params[2])
            log_sigma += 20*dnest4.randh()
            # Note the difference between dnest4.wrap in Python and
            # DNest4::wrap in C++. The former *returns* the wrapped value.
            log_sigma = dnest4.wrap(log_sigma, -10.0, 10.0)
            params[2] = np.exp(log_sigma)

@@

    def log_likelihood(self, params):
        """
        Gaussian sampling distribution.
        """
        m, b, sigma = params
        var = sigma**2
        return -0.5*data.shape[0]*np.log(2*np.pi*var)\
                - 0.5*np.sum((data[:,1] - (m*data[:,0] + b))**2)/var

@@

# Set up the sampler. The first argument is max_num_levels
gen = sampler.sample(max_num_levels=30, num_steps=1000, new_level_interval=10000,
                      num_per_step=10000, thread_steps=100,
                      num_particles=5, lam=10, beta=100, seed=1234)

>>>>

\pkg{MultiNest} \citep{feroz2009multinest} is a
\proglang{Fortran}
Nested Sampling implementation
that dispenses with MCMC for generating new particles from constrained
prior distributions. Instead, the new particles are generated by making
an approximation to the constrained prior using ellipsoids. This tends to work
well in low to moderate dimensional parameter spaces (up to $\sim 40$).
In higher dimensions, \pkg{DNest4} is more useful than \pkg{MultiNest}
since it is based on MCMC. We expect \pkg{MultiNest} to be more useful
than
\pkg{DNest4} only on low to moderate dimensional problems with slow
likelihood evaluations, where \pkg{DNest4}'s reliance on MCMC becomes costly.
The more recent \pkg{POLYCHORD} \citep{handley2015polychord} combines
\pkg{MultiNest}-like methods with MCMC.

@@

\subsection{Generating from the prior}
The member function used to generate straight line parameters from the
prior is:
\begin{CodeChunk}
\begin{CodeInput}
void StraightLine::from_prior(DNest4::RNG& rng)
{
   // Naive diffuse prior
   m = 1E3 * rng.randn();
   b = 1E3 * rng.randn();

   // Log-uniform prior
   sigma = exp( - 10.0 + 20.0 * rng.rand());
}
\end{CodeInput}
\end{CodeChunk}

@@

    if(which == 0)
    {
        // log_H takes care of the prior ratio
        // i.e., log_H = log(pi(theta')/pi(theta))
        log_H -= -0.5 * pow(m / 1E3, 2);

        // Take a step
        m += 1E3 * rng.randh();

        // log_H takes care of the prior ratio
        log_H += -0.5 * pow(m / 1E3, 2);
    }
    else if(which == 1)
    {
        log_H -= -0.5*pow(b / 1E3, 2);
        b += 1E3 * rng.randh();
        log_H += -0.5*pow(b / 1E3, 2);
    }
    else
    {
        // Proposal for a parameter with a log-uniform
        // prior takes the log of the parameter,
        // takes a step with respect to a uniform prior,
        // then takes the exp of the parameter
        sigma = log(sigma);
        sigma += 20.0 * rng.randh();

@@

\begin{CodeChunk}
\begin{CodeInput}
m += 1E3 * rng.randh();
\end{CodeInput}
\end{CodeChunk}

@@

In the example, the proposal for $\sigma$ is implemented by taking advantage
of the uniform prior for $\ln(\sigma)$. So $\sigma$ is transformed by
taking a logarithm, a proposal move is made (that satisfies detailed balance
with respect to a uniform distribution), and then $\sigma$ is exponentiated
again. The step for the uniform prior between -10 and +10
uses the following code:
\begin{CodeChunk}
\begin{CodeInput}
sigma += 20.0 * rng.randh();
DNest4::wrap(sigma, -10.0, 10.0);
\end{CodeInput}
\end{CodeChunk}
The factor of 20 accounts for the prior width, and
the \code{wrap(double\&, double, double)} function uses the modulo operator
to construct periodic boundaries. For example, if the perturbation results
in a value of 10.2, which is outside the prior range, the value is modified to
$-9.8$.
The \code{wrap} function has no return value and works by modifying its
first argument, which is passed by reference.
Alternatively, the log-uniform prior, which has density proportional
to $1/\sigma$, could have been used directly by adding
$\ln\left[(1/\sigma')/(1/\sigma)\right]$ to the
return value instead of using the log/exp trick.
However, this isn't recommended for a prior distribution like this
which covers several orders of magnitude, because the appropriate scale size
for the proposal is less clear. This is likely to cause inefficient
sampling.

@@

    // Variance
    double var = sigma * sigma;

    // Conventional gaussian sampling distribution
    double log_L = 0.0;
    double mu;
    for(size_t i = 0; i < y.size(); ++i)
    {
        mu = m * x[i] + b;
        log_L += -0.5 * log(2 * M_PI * var) - 0.5 * pow(y[i] - mu, 2) / var;
    }

@@

\subsection{Parameter output}
The print and description functions are very simple:
\begin{CodeChunk}
\begin{CodeInput}
void StraightLine::print(std::ostream& out) const
{
    out << m << ' ' << b << ' ' << sigma;
}

@@

    if(which == 0)
    {
        mu += 20 * rng.randh();
        wrap(mu, -10.0, 10.0);
    }
    if(which == 1)
    {
        log_sigma += 20 * rng.randh();
        wrap(log_sigma, -10.0, 10.0);
    }
    if(which == 2)
    {
        int i = rng.rand_int(n.size());
        n[i] = rng.randn();
    }

@@

    // Assemble fake dataset
    vector<double> x_fake = n;
    for(size_t i = 0; i < x_fake.size(); i++)
        x_fake[i] = mu + sigma * x_fake[i];

@@

\subsection{ABC example results}
For ABC applications, there is an alternate \code{postprocess} function
called \code{postprocess_abc}.
This generates posterior samples using the $\epsilon$ value that
compresses the joint prior by 80\% as much as the final level.
\begin{CodeChunk}
\begin{CodeInput}
dnest4.postprocess_abc()
\end{CodeInput}
\end{CodeChunk}
To try different values of $\epsilon$, you can set the argument
\code{threshold\_fraction} to a value other than 0.8. For example
\begin{CodeChunk}
\begin{CodeInput}
dnest4.postprocess_abc(threshold_fraction = 0.6)
\end{CodeInput}
\end{CodeChunk}

@@

    def from_prior(self):
        """
        Unlike in C++, this must *return* a numpy array of parameters.
        """
        m = 1E3 * rng.randn()
        b = 1E3 * rng.randn()
        sigma = np.exp(-10.0 + 20.0 * rng.rand())
        return np.array([m, b, sigma])

@@

        if which == 0 or which == 1:
            logH -= -0.5 * (params[which] / 1E3) **2
            params[which] += 1E3 * dnest4.randh()
            logH += -0.5 * (params[which]/1E3) **2
        else:
            log_sigma = np.log(params[2])
            log_sigma += 20 * dnest4.randh()
            # Note the difference between dnest4.wrap in Python and
            # DNest4::wrap in C++. The former *returns* the wrapped value.
            log_sigma = dnest4.wrap(log_sigma, -10.0, 10.0)
            params[2] = np.exp(log_sigma)

@@

    def log_likelihood(self, params):
        """
        Gaussian sampling distribution.
        """
        m, b, sigma = params
        var = sigma ** 2
        return -0.5 * data.shape[0] * np.log(2 * np.pi * var)\
                - 0.5 * np.sum((data[:,1] - (m * data[:,0] + b)) **2) / var

@@

# Set up the sampler. The first argument is max_num_levels
gen = sampler.sample(max_num_levels=30, num_steps=1000,\
                      new_level_interval=10000,\
                      num_per_step=10000, thread_steps=100,\
                      num_particles=5, lam=10, beta=100, seed=1234)
